### 2015-02-18 doc/thesis/typedlua.tex
Updating syntax documentation
Minor improvements to the text about the informal overview of Typed Lua

In this chapter we use some examples of Typed Lua code to show how
they relate to Lua.
These examples shall give an informal overview of our optional type system.
We will use typing rules to present the formalization of our optional type
system in the next chapter.
All the examples that we present along this chapter run in our Typed Lua compiler.

@@

Lua values can have one of eight tags:
\emph{nil}, \emph{boolean}, \emph{number}, \emph{string},
\emph{function}, \emph{table}, \emph{userdata}, and \emph{thread}.
Typed Lua includes types for the first six.
Typed Lua also includes a syntactical extension that programmers can use
to define the types of \emph{userdata}.
We use this syntactical extension to define the type \emph{thread}. 
In this section we present the Typed Lua types that may appear on annotations.
We explain all Typed Lua types and syntactical extensions along this chapter.

@@

Typed Lua allows optional type annotations in variable and function
declarations.
We use the following example to illustrate how we can annotate a
function declaration and a variable declaration:
\begin{verbatim}
    local function succ (x:integer):integer
      return x + 1
    end
    local x:integer = 7
    x = succ(x)
    print(x)      --> 8
\end{verbatim}

Typed Lua uses local type inference to assign more specific types to
some unannotated declarations.
More precisely, Typed Lua can infer the type of local variables and
the return type of local functions that are not recursive.
The inference that we implement in Typed Lua is quite simple, as it
uses only the type of the local expression.
For local variables, Typed Lua uses the type of the initialization
expression to assign a more specific type to an unannotated local variable.
For local functions, Typed Lua uses the type of the returned expression
to assign a more specific type to an uannotated return type.
This means that we can rewrite the previous example as follows:
\begin{verbatim}
    local function succ (x:integer)
      return x + 1
    end
    local x = 7
    x = succ(x)
    print(x)      --> 8
\end{verbatim}

@@

Typed Lua assigns the type \texttt{any} to the unannotated declarations
that it cannot infer a more specific type.
More precisely, Typed Lua cannot infer more specific types to input
parameters of function declarations and to the return type of
recursive functions, as global type inference is undecidable in a
language with subtyping \citep{wells1999typability}.

@@

Lua has first-class functions, but they have some peculiarities.
First, the number of arguments of a function call does
not need to match the arity of the function declaration, as Lua silently
drops extra arguments after evaluating them, or uses \texttt{nil} to
replace missing arguments.
Second, functions can return any number of values,
and this number of returned values may not be statically known.
Third, Lua has multiple assignment, and the semantics
of argument passing is the same of the multiple assignment, that is,
calling a function is like doing a multiple assignment where the
left hand side is the parameter list and the right hand side is the
argument list.

@@

Typed Lua provides two operation modes: the \emph{default} mode and
the \emph{strict} mode.
In the default mode, the compiler adds a variadic tail to the
type of the parameter list and to the return type whenever the
programmer does not specify one.
The default mode has this behavior to preserve the semantics
of function calls in Lua, that is, it discards extra arguments and
uses \texttt{nil} to replace missing arguments.
In the strict mode, the compiler does not add a variadic tail
to the type of the parameter list and to the return type.
Instead, it preserves the type defined by the programmer to catch
arity mismatch.
We will use the following function to illustrate how these two
operation modes work:
\begin{verbatim}
    local function sum (x:integer, y:integer):integer
      return x + y
    end
\end{verbatim}

@@

As an example,
\begin{verbatim}
    local x:integer, y:integer, z:string = m(), m()
\end{verbatim}
compiles without warnings in both operation modes.
This happens because in the right hand side of the multiple
assignment, only the first value produced by the first call
to \texttt{m} gets used, so the type of the right hand side
is \texttt{(integer, integer, string, nil*)}, which is
consistent with the type of the left hand side
\texttt{(integer, integer, string, value*)}.

@@

Optional values are unions of some type \texttt{t} and \texttt{nil},
and Typed Lua includes the syntactic sugar \texttt{t?} to represent
them because they appear quite often.
The concrete syntax \texttt{t?} is syntactic sugar for \texttt{t|nil}.
Optional values can appear when a function has optional parameters
and when the program reads a value from an array or a map.
The following example shows a function that has an optional parameter:
\begin{verbatim}
    local function message (name:string, greeting:string?)
      local greeting = greeting or "Hello "
      return greeting .. name
    end

    print(message("Typed Lua"))         --> Typed Lua
    print(message("Typed Lua", "Hi "))  --> Hi Typed Lua
\end{verbatim}

Although the parameter \texttt{greeting} is optional, and has type
\texttt{string|nil}, the concatenation does not generate a warning
because we used the short-circuiting \texttt{or} operator to declare a
new variable \texttt{greeting} that is guaranteed to have type \texttt{string}. 
In Lua, only the values \texttt{nil} and \texttt{false} represent a
false condition, so programmers often use the \texttt{or} operator as a
common idiom to assign a default value to an optional parameter. 
Typed Lua uses the following rule to type this idiom: 
if the left hand side of \texttt{or} has type \texttt{t|nil} and
the right hand side has type \texttt{t} then the \texttt{or} expression has
type \texttt{t}.

@@

Typed Lua allows the assignment \texttt{v = v or e} to refine
the type of \texttt{v} from \texttt{t|nil} to \texttt{t}
if it matches the following rules: 
the type of \texttt{e} is a subtype of \texttt{t},
and the variable \texttt{v} is local to the current function
and it is not being assigned in another function.
The refinement only affects the type of \texttt{v} until the
next assignment to \texttt{v}.
In the case of \texttt{greeting}, the assignment refines its type
from \texttt{string|nil} to \texttt{string}.

@@

When Typed Lua finds a union of tuples in the right hand side of
a declaration, it assigns \emph{projection types} to the
variables that appear unannotated in the left hand side of the
declaration.
Projection types do not appear in type annotations, but Typed Lua
uses them to project unions of tuple types into unions of
first-level types that have a dependency relation.
We will discuss projection types in more detail in the next chapter.

@@

The overloading mechanism of Typed Lua have a limitation:
the return type cannot depend on the input type.
Next, we show an example:
\begin{verbatim}
    local function limitation (x:number|string)
      if type(x) == "number" then
        return x + x
      else
        return x .. x
      end
    end
\end{verbatim}

@@

Typed Lua tags a variable that holds a table type as either
\emph{unique}, \emph{open}, or \emph{closed}.
If a variable gets is type from a table constructor then it is
unique, otherwise it is closed.
Any alias to unique variables makes them open, as we use the
tag open to keep track of unique variables that are aliased.
In the previous example, \texttt{person} has an unique type
because it has no alias.
Typed Lua also has different subtyping rules that handle these
different tags.
We explain these tags and their subtyping rules in more detail
in Chapter \ref{chap:system}.

@@

\begin{figure}[!ht]
\textbf{Interfaces}\\
\dstart
\begin{align*}
\textit{interface} ::= & \;\; [\textbf{local}] \; \textbf{interface} \; \textit{typedec}\\
\textit{typedec} ::= & \;\; \textit{Name} \; {\textit{decitem}} \; \textbf{end}\\
\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
\textit{idtype} ::= & \;\; \textit{type} \; | \; \textit{methodtype}\\
\textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
\textit{id} ::= & \;\; [\textbf{const}] \; \textit{Name}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua interfaces}
\label{fig:interfaces}
\end{figure}

@@

Now we can write a program that declares a list, inserts some
elements in this list, and then traverses the list printing
each element of the list:
\begin{verbatim}
    local l:Element?

@@

Note that the type of \texttt{l} is \texttt{Element} in the
first line of the \texttt{while} body, but it is restored to
\texttt{Element|nil} in the second line of the \texttt{while} body
due to the assignment.
These rules allow this example to compile without any warnings,
and we discuss them in more detail in the next chapter.

@@

The type \texttt{Exp} is a recursive type that resembles the
algebraic data types from functional programming.
The small set of \texttt{type} predicates that we mentioned
in Section \ref{sec:unions} also includes a specific rule
that is not based on tags.
This specific rule let programmers discriminate unions of table types
and resembles the pattern matching from functional programming.
We can use this feature to write an evaluation function for
our simple language:
\begin{verbatim}
    local function eval (e:Exp):number
      if e.tag == "Number" then return e[1]
      elseif e.tag == "Add" then return eval(e[1]) + eval(e[2])
      elseif e.tag == "Sub" then return eval(e[1]) - eval(e[2])
      elseif e.tag == "Mul" then return eval(e[1]) * eval(e[2])
      elseif e.tag == "Div" then return eval(e[1]) / eval(e[2])
      end
    end
\end{verbatim}

When Typed Lua finds the predicate $\texttt{v[e}_{\texttt{1}}\texttt{] == e}_{\texttt{2}}$,
it checks whether the type of the variable \texttt{v} is an union of table types,
whether the type of $\texttt{e}_{\texttt{1}}$ is a literal type $\texttt{l}_{\texttt{1}}$,
and whether the type of $\texttt{e}_{\texttt{2}}$ is a literal type $\texttt{l}_{\texttt{2}}$.
If that is the case, it constrains the type of \texttt{v} inside
the \texttt{if} to the table type that has the key type
$\texttt{l}_{\texttt{1}}$ that maps to the type $\texttt{l}_{\texttt{2}}$.
In our example, the expression \texttt{e.tag == "Add"} makes Typed Lua
constrain the type of \texttt{e} to \texttt{\{"tag":"Add", 1:Exp, 2:Exp\}}.
If we did not add this special predicate, we would not be able to traverse the AST,
as any attempt to index an union raises a compile-time warning.

@@

An idiomatic way for defining modules in Lua is to declare only locals
in the source file and return a table constructor at the end of the source file.
The returned constructor includes the members that the module should export.
The following example illustrates this case in Typed Lua:
\begin{verbatim}
    local RADIANS_PER_DEGREE = 3.14159 / 180.0
    local function deg (x:number)
      return x / RADIANS_PER_DEGREE
    end
    local function rad (x:number)
      return x * RADIANS_PER_DEGREE
    end
    local function pow (x:number, y:number)
      return x ^ y
    end
    return {
      deg = deg,
      rad = rad,
      pow = pow,
    }
\end{verbatim}

@@

%There is still a third way for defining modules in Lua through the
%\texttt{module} function, which has been deprecated.
%Even though this style has been deprecated, our survey from
%Section \ref{sec:statistics} indicated that around a third of the
%Lua modules that are in LuaRocks still use this style.
%Thus, Typed Lua also supports this style.

@@

This example shows that Typed Lua assigns the type \texttt{self}
to the implicit parameter \texttt{self}.
The type \texttt{self} represents the type of the \emph{receiver}
in method definitions and method calls.
In this example, the refinement of table types allows Typed Lua
to bind the type \texttt{self} with the same type of \texttt{Shape}:
\begin{verbatim}
    { "x":number, "y":number,
      "move":(self, number, number) -> () }
\end{verbatim}

@@

Furthermore, Typed Lua also allows programmers to create description
files for exporting statically typed interfaces for a Lua modules
that are written in C.

\begin{figure}[!ht]
\textbf{The complete syntax of description files}\\
\dstart
\begin{align*}
\textit{description} ::= & \;\; \textit{desclist}\\
\textit{desclist} ::= & \;\; \textit{descitem} \; \{ \textit{descitem} \}\\
\textit{descitem} ::= & \;\; \textit{typedid} \; | \;
  \textit{interface} \; | \;
  \textit{userdata} \; | \;
  \textit{typealias}\\
\textit{typedid} ::= & \;\; [\textbf{const}] \; \textit{id} \; \texttt{`:'} \; \textit{type}\\ 
\textit{interface} ::= & \;\; \textbf{interface} \; \textit{typedec}\\
\textit{userdata} ::= & \;\; \textbf{userdata} \; \textit{typedec}\\
\textit{typedec} ::= & \;\; \textit{Name} \; {\textit{decitem}} \; \textbf{end}\\
\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
\textit{idtype} ::= & \;\; \textit{type} \; | \; \textit{methodtype}\\
\textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
\textit{id} ::= & \; [\textbf{const}] \; \textit{Name}\\
\textit{typealias} ::= & \;\; \textbf{typealias} \; \textit{Name} \; \texttt{`='} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua description files}
\label{fig:tldsyntax}
\end{figure}

Figure \ref{fig:tldsyntax} shows the complete syntax of Typed Lua
description files in extended BNF.
A Typed Lua description file defines the table type that represents
a certain module and the type names that are exported
along with this table type.
We can export a type name through the declaration of either an interface,
an userdata, or a type alias.
As we mentioned in Section \ref{sec:alias}, a type alias declaration
creates an alias to a more general type, while an interface declaration
creates an alias to a table type that represents a record.
An userdata declaration is similar to an interface declaration,
but it also includes its name as the \emph{brand} of the table type.
The Typed Lua compiler uses this brand to combine structural with
nominal type checking, so two userdata that export exactly the same members,
are not subtype of each other, because they do not share the same brand.

>>>>

In this chapter we use some examples of Typed Lua code to show how
they relate to Lua.
These examples give an informal overview of our optional type system.
We will use typing rules to present the formalization of our optional type
system in the next chapter.
All the examples that we present in this chapter run in our Typed Lua compiler.

@@

Lua values can have one of eight tags:
\emph{nil}, \emph{boolean}, \emph{number}, \emph{string},
\emph{function}, \emph{table}, \emph{userdata}, and \emph{thread}.
Typed Lua includes types for the first six.
Typed Lua also includes a syntactical extension that programmers can use
to define the types of \emph{userdata}.
We use this syntactical extension to define the type \emph{thread}. 
In this section we present the Typed Lua types that may appear on annotations.
We explain all Typed Lua types and syntactical extensions in this chapter.

@@

Typed Lua allows optional type annotations in variable and function
declarations.
We use the following example to illustrate how we can annotate a
function declaration and a variable declaration:
\begin{verbatim}
    local function succ (n:integer):integer
      return n + 1
    end
    local x:integer = 7
    x = succ(x)
    print(x)      --> 8
\end{verbatim}

Typed Lua uses local type inference to assign more specific types to
some unannotated declarations.
More precisely, Typed Lua can infer the type of local variables and
the return type of local functions that are not recursive.
The inference that we implement in Typed Lua is quite simple, as it
uses only the type of the local expression.
For local variables, Typed Lua uses the type of the initialization
expression to assign a more specific type to an unannotated local variable.
For local functions, Typed Lua uses the type of the returned expression
to assign a more specific type to an uannotated return type.
This means that we can rewrite the previous example as follows:
\begin{verbatim}
    local function succ (n:integer)
      return n + 1
    end
    local x = 7
    x = succ(x)
    print(x)      --> 8
\end{verbatim}

@@

Typed Lua assigns the type \texttt{any} to the unannotated declarations
that it does not infer a more specific type.
More precisely, Typed Lua does not infer more specific types to input
parameters of function declarations and to the return type of
recursive functions.

@@

Lua has first-class functions, but they have some peculiarities.
First, the number of arguments of a function call does
not need to match the arity of the function declaration, as Lua silently
drops extra arguments after evaluating them, or uses \texttt{nil} to
replace missing arguments.
Second, functions can return any number of values,
and this number of returned values may not be statically known.
Third, Lua has multiple assignment, and the semantics
of argument passing is the same of the multiple assignment, that is,
calling a function is like doing a multiple assignment where the
left-hand side is the parameter list and the right-hand side is the
argument list.

@@

Typed Lua provides two operation modes: the \emph{default} mode and
the \emph{strict} mode.
In the default mode, the compiler adds a variadic tail to the
type of the parameter list and to the return type if the
programmer does not specify one.
The default mode has this behavior to preserve the semantics
of function calls in Lua, that is, it discards extra arguments and
uses \texttt{nil} to replace missing arguments.
In the strict mode, the compiler does not add a variadic tail
to the type of the parameter list and to the return type.
Instead, it preserves the type defined by the programmer to catch
arity mismatch.

We will use the following function to illustrate how these two
operation modes work:
\begin{verbatim}
    local function sum (x:integer, y:integer):integer
      return x + y
    end
\end{verbatim}

@@

As an example,
\begin{verbatim}
    local x:integer, y:integer, z:string = m(), m()
\end{verbatim}
compiles without warnings in both operation modes.
This happens because in the right-hand side of the multiple
assignment, only the first value produced by the first call
to \texttt{m} gets used, so the type of the right-hand side
is \texttt{(integer, integer, string, nil*)}, which is
consistent with the type of the left-hand side
\texttt{(integer, integer, string, value*)}.

@@

Optional values are unions of some type \texttt{t} and \texttt{nil},
and Typed Lua includes the syntactic sugar \texttt{t?} to represent
them because they appear quite often.
The concrete syntax \texttt{t?} is syntactic sugar for \texttt{t|nil}.
Optional values can appear when a function has optional parameters
and when the program reads a value from an array or a map.
The following example shows a function that has an optional parameter:
\begin{verbatim}
    local function message (name:string, greeting:string?)
      local greeting = greeting or "Hello "
      return greeting .. name
    end
\end{verbatim}

Although the parameter \texttt{greeting} is optional, and has type
\texttt{string|nil}, the concatenation does not generate a warning
because we used the short-circuiting \texttt{or} operator to declare a
new variable \texttt{greeting} that is guaranteed to have type \texttt{string}. 
In Lua, only the values \texttt{nil} and \texttt{false} represent a
false condition, so programmers often use the \texttt{or} operator as a
common idiom to assign a default value to an optional parameter. 
Typed Lua uses the following rule to type this idiom: 
if the left-hand side of \texttt{or} has type \texttt{t|nil} and
the right-hand side has type \texttt{t} then the \texttt{or} expression has
type \texttt{t}.

@@

Typed Lua allows the assignment \texttt{v = v or e} to refine
the type of \texttt{v} from \texttt{t|nil} to \texttt{t}
if it matches the following rules: 
the type of \texttt{e} is a subtype of \texttt{t},
and the variable \texttt{v} is local to the current function
and it is not being assigned in another function.
The refinement only affects the type of \texttt{v} until the
next assignment to \texttt{v}.
In the case of \texttt{greeting}, the assignment constrains its type
from \texttt{string|nil} to \texttt{string}.

After we define the function \texttt{message}, we can call it
with a missing argument in both operation modes:
\begin{verbatim}
    print(message("Typed Lua"))         --> Hello Typed Lua
    print(message("Typed Lua", "Hi "))  --> Hi Typed Lua
\end{verbatim}

The type of \texttt{message} is
\texttt{(string, string|nil, value*) -> (string, nil*)}
in the default mode, and
\texttt{(string, string|nil) -> (string)}
in the strict mode.
The call \texttt{message("Typed Lua")} compiles without
any warnings in both modes, because the argument list
\texttt{("Typed Lua", nil*)} is consistent with
\texttt{(string, string|nil, value*)} in the default mode,
and it is also consistent with  
\texttt{(string, string|nil)} in the strict mode.
The compiler includes \texttt{nil*} in the tail of
an argument list in both modes, as it is necessary
to replace any optional parameters that may appear in
a function declaration.

@@

When Typed Lua finds a union of tuples in the right-hand side of
a declaration, it assigns \emph{projection types} to the
variables that appear unannotated in the left-hand side of the
declaration.
Projection types do not appear in type annotations, but Typed Lua
uses them to project unions of tuple types into unions of
first-level types that have a dependency relation.
We will discuss projection types in more detail in Section \ref{sec:fap}.

@@

The overloading mechanism of Typed Lua has a limitation:
the return type cannot depend on the input types.
Next, we show an example:
\begin{verbatim}
    local function limitation (x:number|string)
      if type(x) == "number" then
        return x + x
      else
        return x .. x
      end
    end
\end{verbatim}

@@

Typed Lua tags a variable that holds a table type as either
\emph{unique}, \emph{open}, \emph{fixed}, or \emph{closed}.
If a variable gets its type from a table constructor then it is
unique, otherwise it is closed.
Any alias to unique variables makes them open, as we use the
tag open to keep track of unique variables that are aliased.
In the previous example, \texttt{person} has an unique type
because it has no alias.
In Section \ref{sec:oop} we will show that we use the tag
fixed to describe the type of classes.
Typed Lua also has different subtyping rules that handle these
different tags.
We explain these tags and their subtyping rules in more detail
in Chapter \ref{chap:system}.

@@

\begin{figure}[!ht]
\textbf{Interfaces}\\
\dstart
\begin{align*}
\textit{interface} ::= & \;\; [\textbf{local}] \; \textbf{interface} \; \textit{typedec}\\
\textit{typedec} ::= & \;\; \textit{Name} \; \{\textit{decitem}\} \; \textbf{end}\\
\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
\textit{idtype} ::= & \;\; \textit{type} \; | \; \textit{methodtype}\\
\textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
\textit{id} ::= & \;\; [\textbf{const}] \; \textit{Name}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua interfaces}
\label{fig:interfaces}
\end{figure}

@@

Now we can write a program that declares a list, inserts some
elements in this list, and then traverses the list printing
each list element:
\begin{verbatim}
    local l:Element?

@@

Note that the type of \texttt{l} is \texttt{Element} inside
the \texttt{while} body, and the assignment \texttt{l = l.next}
restores the type of \texttt{l} to \texttt{Element|nil}.
This means that this example compiles without any warnings, because the
\texttt{while} statement also uses the small set of \texttt{type}
predicates that we mentioned in Section \ref{sec:unions}.

@@

The type \texttt{Exp} is a recursive type that resembles the
algebraic data types from functional programming.
The small set of \texttt{type} predicates that we mentioned
in Section \ref{sec:unions} also includes a specific rule
that is not based on tags.
This specific rule lets programmers discriminate unions of table types
and resembles the pattern matching from functional programming.
We can use this feature to write an evaluation function for
our simple language:
\begin{verbatim}
    local function eval (e:Exp):number
      if e.tag == "Number" then return e[1]
      elseif e.tag == "Add" then return eval(e[1]) + eval(e[2])
      elseif e.tag == "Sub" then return eval(e[1]) - eval(e[2])
      elseif e.tag == "Mul" then return eval(e[1]) * eval(e[2])
      elseif e.tag == "Div" then return eval(e[1]) / eval(e[2])
      end
    end
\end{verbatim}

When Typed Lua finds the predicate $\texttt{v[e}_{\texttt{1}}\texttt{] == e}_{\texttt{2}}$,
it checks whether the type of the variable \texttt{v} is an union of table types,
whether the type of $\texttt{e}_{\texttt{1}}$ is a literal type $\texttt{l}_{\texttt{1}}$,
and whether the type of $\texttt{e}_{\texttt{2}}$ is a literal type $\texttt{l}_{\texttt{2}}$.
If that is the case, it constrains the type of \texttt{v} inside
the \texttt{if} to the table type that has a key of type
$\texttt{l}_{\texttt{1}}$ which maps to a value of type $\texttt{l}_{\texttt{2}}$.
In our example, the expression \texttt{e.tag == "Add"} makes Typed Lua
constrain the type of \texttt{e} to \texttt{\{"tag":"Add", 1:Exp, 2:Exp\}}.
If we did not add this special predicate, we would not be able to traverse an AST,
as any attempt to index an union raises a compile-time warning.

@@

An idiomatic way for defining modules in Lua is to declare only locals
and return a table constructor at the end of the source file.
The returned constructor includes the members that the module should export.
The following example illustrates this case in Typed Lua:
\begin{verbatim}
    local RADIANS_PER_DEGREE = 3.14159 / 180.0
    local function deg (x:number)
      return x / RADIANS_PER_DEGREE
    end
    local function rad (x:number)
      return x * RADIANS_PER_DEGREE
    end
    local function pow (x:number, y:number)
      return x ^ y
    end
    return {
      deg = deg,
      rad = rad,
      pow = pow,
    }
\end{verbatim}

@@

@@

Typed Lua assigns the type \texttt{self} to the implicit parameter \texttt{self}.
The type \texttt{self} represents the type of the \emph{receiver}
in method definitions and method calls.
In this example, Typed Lua binds the type \texttt{self} to
the table type \texttt{\{"x":number, "y":number\}} inside \texttt{move}.
This example also uses the refinement of table types to build the
type of \texttt{Shape}:
\begin{verbatim}
    { "x":number, "y":number,
      "move":(self, number, number) -> () }
\end{verbatim}

@@

Furthermore, Typed Lua also allows programmers to create description
files for exporting statically typed interfaces for Lua modules
that are written in C.

\begin{figure}[!ht]
\textbf{The complete syntax of description files}\\
\dstart
\begin{align*}
\textit{description} ::= & \;\; \textit{desclist}\\
\textit{desclist} ::= & \;\; \textit{descitem} \; \{ \textit{descitem} \}\\
\textit{descitem} ::= & \;\; \textit{typedid} \; | \;
  \textit{interface} \; | \;
  \textit{userdata} \; | \;
  \textit{typealias}\\
\textit{typedid} ::= & \;\; [\textbf{const}] \; \textit{id} \; \texttt{`:'} \; \textit{type}\\ 
\textit{interface} ::= & \;\; \textbf{interface} \; \textit{typedec}\\
\textit{userdata} ::= & \;\; \textbf{userdata} \; \textit{typedec}\\
\textit{typedec} ::= & \;\; \textit{Name} \; \{\textit{decitem}\} \; \textbf{end}\\
\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
\textit{idtype} ::= & \;\; \textit{type} \; | \; \textit{methodtype}\\
\textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
\textit{id} ::= & \;\; [\textbf{const}] \; \textit{Name}\\
\textit{typealias} ::= & \;\; \textbf{typealias} \; \textit{Name} \; \texttt{`='} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua description files}
\label{fig:tldsyntax}
\end{figure}

Figure \ref{fig:tldsyntax} shows the complete syntax of Typed Lua
description files in extended BNF.
A Typed Lua description file defines the table type that represents
a certain module and the type names that are exported
along with this table type.
We can export a type name through the declaration of either an interface,
an userdata, or a type alias.
As we mentioned in Section \ref{sec:alias}, a type alias declaration
creates an alias to a more general type, while an interface declaration
creates an alias to a table type that represents a record.
An userdata declaration is similar to an interface declaration,
but it also includes its name as the \emph{brand} of the table type.
The Typed Lua compiler uses this brand to combine structural with
nominal type checking, so two userdata that export exactly the same members,
but do not have the same name, are not subtype of each other,
because they do not share the same brand.
